---
title: "no2, crossvalidation"
description: |
author:
  - name: guidofioravanti
date: 06-22-2021
output:
  distill::distill_article:
    self_contained: true
---

Cross-validation strategy, **comparing predicted versus observed values: $log(2020/2019)$**

For the model validation, we followed the same approach described in Fioravanti et al. (2020). Namely:

- for each month (January - April), we stratified the input dataset according to the **environment type classification** which distinguishes three different station groups: urban, suburban and rural;

- within each group, 10% of the stations were randomly sampled in order to define a validation dataset. The remainder of the stations (training dataset) was used to fit the model and the fitted model was used to predict the log(2020/2019) on the validation dataset. Comparing predicted and observed values allows to assess the model performance on unseen data (if the model underfits, overfits or generalizes well).

- both the sampling and the estimation process were repeated **three times** for each month. 

### Notation

Months (mm in the results)

1: January; 2: February; 3: March; 4: April

Type site (tipo_zona in the results)

U: urban; S: suburban; R: rural

Model stage (stage in the results)

Val: validation dataset; Train: training dataset

### Results

```{r setup,include=FALSE,echo=FALSE,message=FALSE,error=FALSE,warning=FALSE}
rm(list=objects())
library("tidyverse")
library("stazioniMonitoraggio")
knitr::opts_chunk$set(include=FALSE,echo=FALSE,message=FALSE,error=FALSE,warning=FALSE)

list.files(pattern="^dati.+RDS$")->ffile

purrr::map_dfr(ffile,.f=function(nomeFile){
  
  readRDS(nomeFile)->dati
  ifelse(grepl("Train",nomeFile),"Train","Val")->dati$stage
  str_extract(str_extract(nomeFile,"_indice[1234]_"),"[[:digit:]]")->dati$trial
  
  str_extract(str_extract(nomeFile,"^dati[[:alpha:]]+[0-9]_"),"[0-9]")->mese
  
  dati  %>%
    mutate(mm=mese,yy=2020,dd=day) #%>%
    #mutate(value=exp(value)-1,fitted=exp(fitted)-1)
  
})->mydf


mydf %>%
  mutate(station_eu_code2=paste0(station_eu_code,stage,trial,mm))->mydf

left_join(mydf,stazioni %>% dplyr::select(-i_surface,-altitudedem,-d_a1,-d_a2,-clc_agricultural))->mydf
```

<br><br>

### Observed vs predicted values, boxplot

```{r,include=FALSE}
mydf %>%
  dplyr::select(station_eu_code,stage,trial,value,fitted,mm,yy,tipo_zona) %>%
  rename(observed=value) %>%
  tidyr::gather(key="tipo_valori",value="value",-tipo_zona,-yy,-mm,-station_eu_code,-trial,-stage)->gather_mydf
```


```{r,include=TRUE,fig.width=10,fig.height=12}
ggplot(data=gather_mydf )+
  geom_boxplot(aes(x=tipo_zona,y=value,fill=tipo_valori))+
  facet_wrap(mm~stage,ncol = 2)+
  theme_bw()
```

<br><br>

### Scatterplot observed vs predicted, training dataset (monthly scatterplot)

```{r,include=TRUE,fig.width=8,fig.asp=1}
ggplot(data=mydf %>% filter(stage=="Train"))+
  geom_point(aes(x=value,y=fitted))+
  geom_abline(intercept=0,slope=1,color="red")+
  facet_wrap(~mm)+
  theme_bw()
```

### Scatterplot observed vs predicted, validation dataset (monthly scatterplot)

```{r,include=TRUE,fig.width=8,fig.asp=1}
ggplot(data=mydf %>% filter(stage=="Val"))+
  geom_point(aes(x=value,y=fitted))+
  geom_abline(intercept=0,slope=1,color="red")+
  facet_wrap(~mm)+
  theme_bw()
```

### Scatterplot observed vs predicted, training dataset (monthly scatterplot per type site)


```{r,include=TRUE,fig.width=8,fig.asp=1}
ggplot(data=mydf %>% filter(stage=="Train"))+
  geom_point(aes(x=value,y=fitted))+
  geom_abline(intercept=0,slope=1,color="red")+
  facet_wrap(mm~tipo_zona)+
  theme_bw()
```

### Scatterplot observed vs predicted, validation dataset (monthly scatterplot per type site)

```{r,include=TRUE,fig.width=8,fig.asp=1}
ggplot(data=mydf %>% filter(stage=="Val"))+
  geom_point(aes(x=value,y=fitted))+
  geom_abline(intercept=0,slope=1,color="red")+
  facet_wrap(mm~tipo_zona)+
  theme_bw()
```

### RMSE (observed vs predicted values)

The boxplots show the distribution of the RMSE values for the training and validation stations.

```{r,include=FALSE}
mydf%>%
  mutate(error=(value-fitted)^2) %>%
  group_by(station_eu_code,mm,stage,tipo_zona) %>%
  summarise(rmse=sqrt(mean(error,na.rm=TRUE))) %>%
  ungroup()->valoriRmse
```

<br><br>

#### RMSE values ditribution per month and stage (training/validation)

```{r,include=TRUE}
print(skimr::skim(valoriRmse %>% dplyr::select(mm,stage,rmse) %>% group_by(mm,stage)))
```

<br><br>

#### RMSE values ditribution per month, type site and stage 

```{r,include=TRUE}
print(skimr::skim(valoriRmse %>% dplyr::select(mm,stage,rmse,tipo_zona) %>% group_by(mm,stage,tipo_zona)))
```

<br><br>

```{r,include=TRUE,fig.height=6,fig.width=12}
ggplot(data=valoriRmse,aes(x=mm,y=rmse))+
  geom_boxplot()+
  facet_wrap(~stage)+
  theme_bw()
```

<br><br>

### Correlation (observed vs predicted values)

The boxplots show the distribution of the correlation values for the training and validation stations.

```{r,include=FALSE}
mydf%>%
  group_by(station_eu_code,mm,stage,tipo_zona) %>%
  summarise(corr=cor(value,fitted,use="pairwise.complete.obs")) %>%
  ungroup()->valoriCor
```

<br><br>

#### Correlation values ditribution per month and stage (training/validation)

```{r,include=TRUE}
print(skimr::skim(valoriCor %>% dplyr::select(mm,stage,corr) %>% group_by(mm,stage)))
```

<br><br>

#### Correlation values ditribution per month, type site (urban, suburban and rural) and stage (training/validation)

```{r,include=TRUE}
print(skimr::skim(valoriCor %>% dplyr::select(mm,stage,corr,tipo_zona) %>% group_by(mm,stage,tipo_zona)))
```

<br><br>

```{r,include=TRUE,fig.height=6,fig.width=12}
ggplot(data=valoriCor,aes(x=mm,y=corr))+
  geom_boxplot()+
  facet_wrap(~stage)+
  theme_bw()
```


```{r,include=TRUE,eval=FALSE}
cor(mydf[mydf$stage=="Val",c("fitted","value")],use = "pairwise.complete.obs")
```


```{r,include=TRUE,eval=FALSE}
cor(mydf[mydf$stage=="Train",c("fitted","value")],use = "pairwise.complete.obs")
```


```{r,include=TRUE,eval=FALSE}
cor(mydf[mydf$tipo_zona=="U" & mydf$stage=="Train",c("fitted","value")],use = "pairwise.complete.obs")
```


```{r,include=TRUE,eval=FALSE}
cor(mydf[mydf$tipo_zona=="U" & mydf$stage=="Val",c("fitted","value")],use = "pairwise.complete.obs")
```


```{r,include=TRUE,eval=FALSE}
cor(mydf[mydf$tipo_zona=="R" & mydf$stage=="Train",c("fitted","value")],use = "pairwise.complete.obs")
```


```{r,include=TRUE,eval=FALSE}
cor(mydf[mydf$tipo_zona=="R" & mydf$stage=="Val",c("fitted","value")],use = "pairwise.complete.obs")
```

```{r,include=TRUE,eval=FALSE}
cor(mydf[mydf$tipo_zona=="S" & mydf$stage=="Train",c("fitted","value")],use = "pairwise.complete.obs")
```

```{r,include=TRUE,eval=FALSE}
cor(mydf[mydf$tipo_zona=="S" & mydf$stage=="Val",c("fitted","value")],use = "pairwise.complete.obs")
```

<br><br>

```{r,include=FALSE}
mydf %>%
  filter(stage=="Train") %>%
  mutate(station_eu_code=paste0(station_eu_code,stage,trial,mm))->training
```


`r sqrt(mean((training$fitted-training$value)^2,na.rm=TRUE))`

`r mean((training$fitted-training$value),na.rm=TRUE)`

```{r,include=FALSE,eval=FALSE}
pdf("training.pdf",width=12,height=8,onefile = TRUE)
purrr::walk(unique(training$station_eu_code),.f=function(codex){
  ggplot(data=training %>% filter(station_eu_code==codex) )+
    geom_line(aes(x=dd,y=value,group=station_eu_code))+
    geom_line(aes(x=dd,y=fitted,group=station_eu_code),color="red")+
    facet_wrap(~yy)+
    theme_bw()->grafico
  
  print(grafico)
  
})
dev.off()
```


```{r,include=FALSE}
mydf %>%
  filter(stage=="Val") %>%
  mutate(station_eu_code=paste0(station_eu_code,stage,trial,mm))->validazione
```

`r sqrt(mean((validazione$fitted-validazione$value)^2,na.rm=TRUE))`

`r mean((validazione$fitted-validazione$value),na.rm=TRUE)`

```{r,include=FALSE,eval=FALSE}
pdf("validazione.pdf",width=12,height=8,onefile = TRUE)
purrr::walk(unique(validazione$station_eu_code),.f=function(codex){

  ggplot(data=validazione %>% filter(station_eu_code==codex) )+
    geom_ribbon(aes(x=dd,ymin=fitted025,ymax=fitted975),fill="lightgray")+
    geom_line(aes(x=dd,y=value,group=station_eu_code))+
    geom_line(aes(x=dd,y=fitted,group=station_eu_code),color="red")+
    facet_wrap(~yy)+
    theme_bw()->grafico
  
  print(grafico)

})
dev.off()
```



